---
title: "esm244_assignment_1_part2_LBleifuss"
author: "Lydia Bleifuss"
date: "2/1/2020"
output: html_document
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(naniar)
library(VIM)
library(skimr)
library(ggfortify)
library(here)
```

```{r}
## 1. Read in the data

env_data <- read_csv(here("Assignment 1 data-20200128","world_env_vars.csv"))

```

```{r}
#2.  Basic cleanign and wrangling before exploration and more cleaning 

env_clean <- env_data %>% 
  clean_names() %>% #makes column headings lower case and with no spaces (_ instead)
   mutate(country = str_to_lower(country)) %>% #county column rows were capitalized so I changed to make coding easier
  
```

```{r}
#3. Explore missingness to get to know which variable may be more descriptive than others and where selection bias may come into play 

summary(env_clean) #Want to see where there are NA values

#Most variables have ~6 NAs, however, wind has 55 our of 243 values missing and therefore will likely want to exclude from the analysis 

gg_miss_var(env_clean) #wanted to check the summary analysis above, using gg_miss_var to which creates a very helpful graph indicating which variables have the most missingness, in this case it is wind (far and away more NAs) 

matrixplot(env_clean) #higher values are darker and red indicates NAs 

matrixplot(env_clean, sortby = "tree_canopy_cover") #sorted by a few speciic variables, organizing variable values from high to low, to see what other values display a similar gradient and therefore may be more correlated 


```


Remove (complete case deletion) any observation with ANY NA values: #using "ANY NA, it will remove any ROW that has a missing variable"

```{r}
#5. Remove NAs now having an understanding for where they are/how many 

env_no_na <- env_clean %>%
  drop_na() %>% 
  select(-wind, -country) #got rid of wind because too many NAs, took out country because PCA needs to have only numeric values, and also 

summary(env_no_na) #now I see that NAs are gone and total observations adjusted accordingly 

skim(env_no_na) #more in depth summary, also tells me that all columns besides character "country" are numeric 
```

```{r, eval = FALSE}
#Need to add "eval - FALSE" to header so R doesn't run when knitting 

#5. PCA Analysis 

#ERROR, good example 
env_pca<- prcomp(env_no_na, scale = TRUE) #normalizing the scales accross units

```


```{r}
my_ca_pca <- prcomp(ca_pb_nona, scale = TRUE)  

my_ca_pca
summary(my_ca_pca) #shows how much (proportion) of the variance is explained by each principle component (PC#)

#biplot(my_ca_pca) #scary, moving on to autoplot

#going to use auto plot which is why we installed fortify 

my_biplot <- autoplot(my_ca_pca,
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE)+
scale_y_continuous(limits = c(-0.05, 0.05))

my_biplot

#this is only representing about 33% of varience, not great but...

#positively correlated: unemployment and poverty 
#no correlation: ozone and diesel_pm
#negatively correlation: drinking_water, groundwater_threats 

#because autoplot works with ggplot, you can use ggplot techniques to start customizing, but, lets look at what variables could be correlated to explor that 
```

## 4. PCA for polluiton burden + demographics 

Join together the ca_dem_clean with ca_pb_nopct and we will drop all of the NAs once we have joined them

```{r}
##you can always use full join to keep everything! Then you can decide what to drop. (MERGE is the SAME, it's just the base r version)

ca_df <- ca_dem_clean %>% 
  inner_join(ca_pb_nopct, by = c("census_tract_number" = "census_tract")) #use c because then you can add a comma and add more, the left of the = corresponds to the left dataframe

#Now drop NAs by complete case deletion:

ca_nona <- ca_df %>% 
  drop_na()

```

Make a subset of ca_nona that only includes: 

- white_percent
- elderly_65_percent
- pm2_5
- pesticides
- traffic
- asthma
- cardiovascular_disease
- poverty 

```{r}
my_sub <- ca_nona %>% 
  select(white_percent,white_percent, elderly_65_percent, pm2_5, pesticides, traffic, asthma, cardiovascular_disease, poverty)
```

PCA:
```{r}
my_dem_pca <- prcomp(my_sub, scale = TRUE)

my_dem_biplot <- autoplot(my_dem_pca,
                      colour = NA,
                      loadings.label = TRUE,
                      loadings.label.size = 3,
                      loadings.label.colour = "black",
                      loadings.label.repel = TRUE)+
scale_y_continuous(limits = c(-0.05, 0.05))+
  theme_minimal()

my_dem_biplot #this explains over 50% (add 36.14% plut 16.04%) so that's nice 
```


```

